---
title: "[Last 200915, 18project, C2C12] MakeCountmatrix_forGEO_201012 CELseq_QC_Iwasaki_0386_C2C12_H3mm18_noumi_20190613_0722_fin191203_200523ver_last200915"
output:
  html_notebook: 
    toc: yes
  pdf_document: 
    keep_tex: yes
    latex_engine: lualatex
---

### Setup

```{r libraries,message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
source("/home/guestA/n70275b/work/rscripts/geomNorm.R")

# Helper function
#ggpoints <- function(x,...) 
#  ggplot(x,...) + geom_point(size=3,stroke=1) +
#  ggrepel::geom_text_repel(size=4) + theme_minimal() + mycolor

## ラベルあり
ggpoints <- function(x,...) 
  ggplot(x,...) + geom_point(stroke=1) +
  ggrepel::geom_text_repel(size=4) + theme_minimal() + mycolor

## ラベルなし
#ggpoints <- function(x,...) 
#  ggplot(x,...) + geom_point(stroke=1) + theme_minimal() + mycolor


print(Sys.Date())
print(sessionInfo(),locale=FALSE)

select <- dplyr::select
rename <- dplyr::rename #191203
count <- dplyr::count #191203



```

### Parameters

*modify here*

```{r params}
# Files

#deftable <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200523/iwasaki_0386_noumi_def_fin191203__200523ver.txt" #最終版 121203
deftable <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/deftable_used_CEL0386noumi_C2C12_fin200915.csv" #最終版 121203


#deftable <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/iwasaki_0386_umi_def_fin191203ver.txt" #最終版 121203
#deftable <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/iwasaki_0386_umi_def.txt"

#deftable <- "deftable_BRB_umi_new.txt"

## Data selection (filter rows of deftable)
#use <- quo(!grepl("^18",group) & (group != "Nc-minusTryd"))
#use <- quo(TRUE) # use all
use <-  quo(group %in% c("eGFP_G","mm18B_G","eGFP_D72", "mm18B_D72"))
#use <- quo(type == "C2C12")
#use <- quo(type != "C2C12")
#use <- quo(TRUE)

#use <- quo(type == "Whole_cell")
#use <- quo(type == "Nucleus")

# Species specific parameters
species <- "Mus musculus"
biomartann <- "mmusculus_gene_ensembl"
maxchrom <- 19 # 19: mouse, 22: human

# Graphics
# aesthetic mapping of labels
#myaes <- aes(colour=enzyme,shape=leg,label=rep) 

#myaes <- aes(colour=growth,shape=type,size=count) #ラベルなし
#myaes <- aes(colour=growth,shape=type,label=replicate,size=count) #ラベルあり
#myaes <- aes(colour=enzyme,shape=leg,label=replicate) #ラベルあり
#myaes <- aes(colour=enzyme,shape=leg,label=factor(rep))

#myaes <- aes(colour=type,shape=trypsin,label=factor(lot)) 
#myaes <- aes(colour=trypsin,label=factor(lot)) 
myaes <- aes(colour=time,shape=type,label=factor(lot)) 

# color palette of points: See vignette("ggsci")
#mycolor <- ggsci::scale_color_aaas()
mycolor <- ggsci::scale_color_d3("category20") # color palette of points

# PCA/UMAP
scalerows <- TRUE # gene-wise scaling (pattern is the matter?)
ntop <- 500 # number of top-n genes with high variance
seed <- 123 # set another number if UMAP looks not good
n_nei <- 2  #6 # number of neighboring data points in UMAP

#hashigushi
scalerows <- FALSE # gene-wise scaling (pattern is the matter?)
#seed <- 123 # set another number if tSNE looks not good
#perprexity <- 3 # expected cluster size in tSNE

# DESeq2
#model <- ~type+trypsin

#model <- ~trypsin
model <- ~group


fdr <- 0.1 # acceptable false discovery rate
lfcthreth <- log2(1) # threshold in abs(log2FC)
# controls should be placed in the right
contrast <- list( 
  
  #time_D72_vs_G = c("time", "D72", "G")
  


  #group_G_H3f3b_vs_WT = c("group", "H3f3b_G", "WT_G"),
  #group_G_mm18B_vs_WT = c("group", "mm18B_G", "WT_G"),
  #group_G_eGFP_vs_WT = c("group", "eGFP_G", "WT_G"),
  #group_G_H3f3b_vs_eGFP = c("group", "H3f3b_G", "eGFP_G"),
  group_G_mm18B_vs_eGFP = c("group", "mm18B_G", "eGFP_G"),
  #group_G_mm18B_vs_H3f3b = c("group", "mm18B_G", "H3f3b_G"),
  
  #group_D72_H3f3b_vs_WT = c("group", "H3f3b_D72", "WT_D72"),
  #group_D72_mm18B_vs_WT = c("group", "mm18B_D72", "WT_D72"),
  #group_D72_eGFP_vs_WT = c("group", "eGFP_D72", "WT_D72"),
  #group_D72_H3f3b_vs_eGFP = c("group", "H3f3b_D72", "eGFP_D72"),
  group_D72_mm18B_vs_eGFP = c("group", "mm18B_D72", "eGFP_D72"),
  #group_D72_mm18B_vs_H3f3b = c("group", "mm18B_D72", "H3f3b_D72")
  
    #group_WT_D72_vs_G = c("group", "WT_D72", "WT_G"),
  group_eGFP_D72_vs_G = c("group", "eGFP_D72", "eGFP_G"),  
  #group_H3f3b_D72_vs_G = c("group", "H3f3b_D72", "H3f3b_G"),
  group_mm18B_D72_vs_G = c("group", "mm18B_D72", "mm18B_G")
  

  
  #type = c("type", "Nucleus", "Whole_cell"),
  #trypsin = c("trypsin", "plus", "untreated")
  
  #Intercept = list("Intercept"), # reference level
  #leg_LvsR = c("leg", "L", "R"),
  #enz_KvsC = c("enzyme","K","C")
  #legL.enzK = list("legL.enzymeK") # interaction
  
  #type_Doxplus_vs_minus = c("type", "Doxplus", "Doxminus")
)
```



### Retrieve Biomart

```{r biomart, cache=TRUE}
#if(!exists("e2g")){
#  ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL",host="asia.ensembl.org")
#  #ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL",host="uswest.ensembl.org")
#  #ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL",host="useast.ensembl.org")
#  mart <- biomaRt::useDataset(biomartann,mart=ensembl)
#  e2g <- biomaRt::getBM(attributes=c("ensembl_gene_id","external_gene_name",
#    "gene_biotype","chromosome_name"), mart=mart) %>% as_tibble %>%
#  rename(
#    ens_gene = ensembl_gene_id,
#    ext_gene = external_gene_name,
#    biotype = gene_biotype,
#    chr = chromosome_name
#  )
#}
#annotate <- partial(right_join,e2g,by="ens_gene")

#-----#
#nrow(e2g)
#readr::write_csv(e2g,"ensemble_list_asia__fin200915.csv")
##readr::write_csv(e2g,"ensemble_list_uswest_fin200523.csv.csv")
###readr::write_csv(e2g,"ensemble_list_useast.csv")

e2g <- readr::read_csv("/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/ensemble_list_asia__fin200915.csv")
annotate <- partial(right_join,e2g,by="ens_gene")
nrow(e2g)

```
### Load counts

```{r loadUMI}
def <- readr::read_csv(deftable) %>% arrange(group,sample) #202010012 change
print(def)
def <- def %>% dplyr::filter(group %in% c("mm18B_G", "eGFP_G","mm18B_D72", "eGFP_D72")) %>% mutate(group=factor(group,c("eGFP_G","mm18B_G","eGFP_D72", "mm18B_D72"))) %>% arrange(group,sample)
print(def)

#def <- readr::read_tsv(deftable) %>% filter(!!use) %>% arrange(group,sample) #20200915 change
#print(def)
#readr::write_csv(def,"deftable_used_CEL0386noumi_C2C12_fin200915.csv")

####--- New ---#### (no UMI ?)
# Set reference levels according to the contrast
for(x in keep(contrast,is.character))
  def[[x[1]]] <- relevel(factor(def[[x[1]]]),x[3])

umi <- def$file %>% unique %>% tibble(file=.) %>% 
  dplyr::mutate(data=map(file,readr::read_tsv,progress=FALSE)) %>%
  unnest() %>% dplyr::rename(barcode=cell) %>%
  dplyr::inner_join(select(def,file,barcode,sample),.,c("file","barcode")) %>%
  select(-file,-barcode) %>% dplyr::rename(ens_gene=gene)

print(umi)

## sample, barcode, file を忘れずに！

mat <- umi %>% annotate %>%
  dplyr::mutate(chr=factor(chr,c(1:maxchrom,"X","Y","MT"))) %>%
  filter(!is.na(chr)) %>% spread(sample,count,fill=0)
## to check read vias, this add read number as "n" column (2019/4/19)
def <- umi %>% count(sample,wt=count) %>% dplyr::inner_join(def,.) %>% dplyr::rename(count=n)
####-----------#### 


#def <- readr::read_tsv(deftable) %>% filter(!!use)

####--- New ---#### (no UMI ?)
# Set reference levels according to the contrast
#for(x in keep(contrast,is.character))
#  def[[x[1]]] <- relevel(factor(def[[x[1]]]),x[3])

#umi <- def$file %>% unique %>% tibble(file=.) %>% 
#  dplyr::mutate(data=map(file,readr::read_tsv,progress=FALSE)) %>%
#  unnest() %>% dplyr::rename(barcode=cell) %>%
#  inner_join(select(def,file,barcode,sample),.,c("file","barcode")) %>%
#  select(-file,-barcode) %>% dplyr::rename(ens_gene=gene)

## sample, barcode, file を忘れずに！

#mat <- umi %>% annotate %>%
#  dplyr::mutate(chr=factor(chr,c(1:maxchrom,"X","Y","MT"))) %>%
#  filter(!is.na(chr)) %>% spread(sample,count,fill=0)
## to check read vias, this add read number as "n" column (2019/4/19)
#def <- umi %>% count(sample,wt=count) %>% inner_join(def,.) %>% dplyr::rename(count=n)
####-----------#### 

# Old
# Set reference levels according to the contrast
#for(x in keep(contrast,is.character))
#  def[[x[1]]] <- relevel(factor(def[[x[1]]]),x[3])
#umi <- def$file %>% unique %>% tibble(file=.) %>% 
#  mutate(data=map(file,readr::read_tsv,progress=FALSE)) %>%
#  unnest() %>% rename(barcode=cell) %>%
#  inner_join(select(def,file,barcode,sample),.,c("file","barcode")) %>%
#  select(-file,-barcode) %>% rename(ens_gene=gene)
#mat <- umi %>% annotate %>%
#  mutate(chr=factor(chr,c(1:maxchrom,"X","Y","MT"))) %>%
#  filter(!is.na(chr)) %>% spread(sample,count,fill=0)



print(umi)
print(def)
```
### Reads breakdown

#### Total reads

```{r totalReads, fig.width=7,fig.height=5}

bychr <- mat %>% select(-(1:3)) %>%
  gather("sample","count",-chr) %>%
  group_by(chr,sample) %>% summarise(total=sum(count)) %>% ungroup

ggplot(bychr,aes(reorder(sample,dplyr::desc(sample)),total/1e6,fill=chr)) +
  theme_linedraw() + geom_bar(stat="identity") + coord_flip() +
  xlab("sample") + ylab("million reads") + ggsci::scale_fill_igv() +
  scale_x_discrete(limits = rev(levels(sample)))


# 前
#bychr <- mat %>% select(-(1:3)) %>%
#  gather("sample","count",-chr) %>%
#  group_by(chr,sample) %>% summarise(total=sum(count)) %>% ungroup

#ggplot(bychr,aes(reorder(sample,desc(sample)),total/1e6,fill=chr)) +
#  theme_linedraw() + geom_bar(stat="identity") + coord_flip() +
#  xlab("sample") + ylab("million reads") + ggsci::scale_fill_igv() +
#  scale_x_discrete(limits = rev(levels(sample)))


#bychr <- mat %>% select(-(1:3)) %>%
#  gather("sample","count",-chr) %>%
#  group_by(chr,sample) %>% summarise(total=sum(count)) %>% ungroup
#ggplot(bychr,aes(reorder(sample,dplyr::desc(sample)),total/1e6,fill=chr)) +
#  theme_linedraw() + geom_bar(stat="identity") + coord_flip() +
#  xlab("sample") + ylab("million reads") + ggsci::scale_fill_igv() +
#  scale_x_discrete(limits = rev(levels(sample)))
```

#### Biotype

```{r biotype,fig.width=8,fig.height=7}
bt <- mat %>% select(-c(1,2,4)) %>% group_by(biotype) %>%
  summarise_all(sum) %>% filter_at(-1,any_vars(. > 1000))
bt %>% tibble::column_to_rownames("biotype") %>%
  as.matrix %>% t %>% mosaicplot(las=2,shade=TRUE)
```

### Correlations

drop rows with all 0 -> +1/2 -> geom.scale -> log -> Pearson's

```{r makemat, fig.width=8,fig.height=7}
matf <- mat %>% filter(chr!="MT") %>% filter_at(-(1:4),any_vars(. > 0))
X <- matf %>% select(-(1:4)) %>% as.matrix
rownames(X) <- matf$ens_gene
lX <- log(gscale(X+0.5))
R <- cor(lX); diag(R) <- NA
pheatmap::pheatmap(R,color=viridis::viridis(256))
```

## Save Matrix (for GEO)

```{r load data setting}
normcount_readfile <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/CEL0386noumi_C2C12_normCount_genename_fin200915.csv"
VST_readfile <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/CEL0386noumi_C2C12_H3mm18__vst_type_all_fin200915.csv"
zscore_readfile <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/CEL0386noumi_C2C12_H3mm18__zscore_type_all_fin200915.csv"
```


### Save Count Matrix (for GEO)
Original Count Matrix

```{r save original Count Matrix}

count_original_writefile1 <- basename(normcount_readfile) %>% gsub("normCount","Count",.) %>% gsub(".csv","__all_CountMatrix_type.csv",.)
count_original_writefile2 <- basename(normcount_readfile) %>% gsub("normCount","Count",.) %>% gsub(".csv","__all_CountMatrix.csv",.)

# MT genes, 0ありのカウントも含む。
nrow(mat)
ncol(mat)
colnames(mat)
countMatrix_original <- mat %>% dplyr::select(ens_gene,ext_gene, biotype, chr, all_of(def$sample))
head(countMatrix_original)
nrow(countMatrix_original)
ncol(countMatrix_original)
colnames(countMatrix_original)
##
print(count_original_writefile1)
countMatrix_original %>% readr::write_csv(count_original_writefile1)
##
countMatrix_original %>% dplyr::select(-ext_gene, -biotype, -chr) %>% head()
print(count_original_writefile2)
countMatrix_original %>% dplyr::select(-ext_gene, -biotype, -chr) %>% readr::write_csv(count_original_writefile2)

```



Count Matrix used in DESeq

```{r save Count Matrix used in DEG}
count_writefile1 <- basename(normcount_readfile) %>% gsub("normCount","Count",.) %>% gsub(".csv","___CountMatrix_type.csv",.)
count_writefile2 <- basename(normcount_readfile) %>% gsub("normCount","Count",.) %>% gsub(".csv","___CountMatrix.csv",.)



# XのデータをDEGに使用 X <- matf %>% select(-(1:4)) %>% as.matrix rownames(X) <- matf$ens_gene #out of 23695 with nonzero total read count
nrow(matf)
ncol(matf)
colnames(matf)

countMatrix <- matf %>% dplyr::select(ens_gene,ext_gene, biotype, chr, all_of(def$sample))

head(countMatrix)
nrow(countMatrix)
ncol(countMatrix)
colnames(countMatrix)

##
print(count_writefile1)
countMatrix %>% readr::write_csv(count_writefile1)
##
countMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% head()
print(count_writefile2)
countMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% readr::write_csv(count_writefile2)


```
### Save normalized count Matrix (for GEO)
```{r load normcount}

normcount <- readr::read_csv(normcount_readfile)


```

```{r save Normalized Count Matrix used in DEG}

normcount_writefile1 <- basename(normcount_readfile) %>% gsub(".csv","___NormalizedCountMatrix_type.csv",.)
normcount_writefile2 <- basename(normcount_readfile) %>% gsub(".csv","___NormalizedCountMatrix.csv",.)

#out of 23695 with nonzero total read count
nrow(normcount)
ncol(normcount)
colnames(normcount)

NormcountMatrix <- normcount %>% dplyr::select(ens_gene,ext_gene, biotype, chr, all_of(def$sample))

head(NormcountMatrix)
nrow(NormcountMatrix)
ncol(NormcountMatrix)
colnames(NormcountMatrix)

##
print(normcount_writefile1)
NormcountMatrix %>% readr::write_csv(normcount_writefile1)
##
NormcountMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% head()
print(normcount_writefile2)
NormcountMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% readr::write_csv(normcount_writefile2)


```

### Save VST Matrix (for GEO)
```{r load vst}

VST <- readr::read_csv(VST_readfile)

```

```{r save VST Matrix used in DEG}
VST_writefile1 <- basename(VST_readfile) %>% gsub(".csv","___VSTMatrix_type.csv",.)
VST_writefile2 <- basename(VST_readfile) %>% gsub(".csv","___VSTMatrix.csv",.)

#out of 23695 with nonzero total read count
nrow(VST)
ncol(VST)
colnames(VST)

VSTMatrix <- VST %>% dplyr::select(ens_gene,ext_gene, biotype, chr, all_of(def$sample))

head(VSTMatrix)
nrow(VSTMatrix)
ncol(VSTMatrix)
colnames(VSTMatrix)

##
print(VST_writefile1)
VSTMatrix %>% readr::write_csv(VST_writefile1)
##
VSTMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% head()
print(VST_writefile2)
VSTMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% readr::write_csv(VST_writefile2)

```


### Save Zscore Matrix (for GEO)
```{r load zscore}

zscore <- readr::read_csv(zscore_readfile)


```

```{r save Zscore Matrix used in DEG}
zscore_writefile1 <- basename(zscore_readfile) %>% gsub(".csv","___ZscoreMatrix_type.csv",.)
zscore_writefile2 <- basename(zscore_readfile) %>% gsub(".csv","___ZscoreMatrix.csv",.)

#out of 23695 with nonzero total read count
nrow(zscore)
ncol(zscore)
colnames(zscore)

ZscoreMatrix <- zscore %>% dplyr::select(ens_gene,ext_gene, biotype, chr, all_of(def$sample))

head(ZscoreMatrix)
nrow(ZscoreMatrix)
ncol(ZscoreMatrix)
colnames(ZscoreMatrix)

##
print(zscore_writefile1)
ZscoreMatrix %>% readr::write_csv(zscore_writefile1)
##
ZscoreMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% head()
print(zscore_writefile2)
ZscoreMatrix %>% dplyr::select(-ext_gene, -biotype, -chr) %>% readr::write_csv(zscore_writefile2)

```

### def

```{r save def}
print(def)

def_writefile1 <- basename(count_original_writefile1) %>% gsub("__all_CountMatrix_type.csv",".csv",.) %>% gsub("_Count_genename","_deftable",.) 
print(def_writefile1)
def %>% readr::write_csv(def_writefile1)
```
def %>% mutate(filename=basename(file)) %>% dplyr::select(sample,barcode,group,type,time)

### log2FC

```{r load log2FC}

#それぞれのlog2FC (読み込みが大変なため、そのまま登録)
filepath_log2FC <- "/home/guestA/o70578a/akuwakado/kuwakado/scCELSeq2/Iwasaki_0386_C2C12_H3mm18/Final_Last_Rserver_200915/2gun/Iwasaki_0386re_C2C12_H3mm18_noumi_resultsall__final200915.csv"
log2FC_writefile2 <- basename(filepath_log2FC) %>% gsub("Iwasaki_0386re_","",.)

####

log2FC <- readr::read_csv(filepath_log2FC)
print(log2FC)
colnames(log2FC)
log2FC %>% group_by(aspect) %>%  summarise(n())

####
log2FC_sel <- log2FC %>% filter(aspect %in% c("group_G_mm18B_vs_eGFP","group_D72_mm18B_vs_eGFP"))
print(log2FC_sel)
colnames(log2FC_sel)
log2FC_sel %>% group_by(aspect) %>%  summarise(n())
print(log2FC_writefile2)
log2FC_sel %>% readr::write_csv(log2FC_writefile2)
```


